---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# Spatial Omic Structure Analysis

<!-- badges: start -->
<!-- badges: end -->

`sosta` (spatial omics structure analysis) is a framework to reconstruct, characterise and compare spatial structures from spatial omics data.

`sosta` builds on existing R packages for spatial analysis such as `spatstat`, `sf` and `SpatialFeatureExperiment`, which facilitates the implementation of additional custom metrics that fit the needs of users.

## Installation

You can install the development version of `sosta` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("sgunz/sosta")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example, eval = FALSE}
library(sosta)
## basic example code
```

