---
title: "Reconstruction of anatomical structures"
author: 
  - name: "Samuel Gunz"
    affiliation:
      - &DMLS Department of Molecular Life Sciences, University of Zurich, Switzerland
      - &SIB SIB Swiss Institute of Bioinformatics, University of Zurich, Switzerland
    email: "samuel.gunz@uzh.ch"
  - name: Mark D. Robinson
    affiliation:
      - *DMLS
      - *SIB
package: "`r BiocStyle::Biocpkg('sosta')`"
output:
  BiocStyle::html_document
abstract: >
  Reconstruction and analysis of pancreatic islets from IMC data using the `sosta` package
vignette: >
  %\VignetteIndexEntry{Reconstruction and analysis of pancreatic islets from IMC data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: sosta.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Installation

`r BiocStyle::Biocpkg('sosta')` can be loaded from Bioconductor and installed as follows

```{r installation, include = TRUE, eval=FALSE}
if (!requireNamespace("BiocManager")) {
    install.packages("BiocManager")
}
BiocManager::install("sosta")
```

# Setup

```{r setup, message=FALSE}
library("sosta")
library("dplyr")
library("tidyr")
library("ggplot2")
library("ggspavis")
library("imcdatasets")
```

# Introduction


```{r loading, echo=FALSE, message=FALSE}
# load the data
data("sostaSPE")
``` 

First we plot the data for illustration. As we have multiple images per patient we will subset one patient and a few slides.

```{r}
plotSpots(
    sostaSPE,
    annotate = "cell_type",
    sample_id = "sample_id",
    in_tissue = NULL,
    y_reverse = FALSE
) + facet_wrap(~image_name)
```

The goal is to reconstruct / segment the structures given by cell type A.

## Reconstruction of structures for one image

In this example we will reconstruct the islets based on the point pattern density of the cells of type A. We will start with estimating the parameters that we use for reconstruction afterwards. For one image this can be illustrated as follows.

```{r}
shapeIntensityImage(
    sostaSPE,
    marks = "cell_type",
    image_col = "image_name",
    image_id = "image1",
    mark_select = "A"
)
```

We see the density (pixel) image on the left and a histogram of the intensity values on the right. The estimated threshold is roughly the mean between the two modes of the (truncated) pixel intensity distribution.

This was done for one image. The function `estimateReconstructionParametersSPE` returns two plots with the estimated parameters for each image. The parameter `bndw` is the bandwidth parameter that is used for estimating the intensity profile of the point pattern. The parameter `thres` is the estimated parameter for the density threshold for reconstruction. We subset 20 images to speed up computation.

```{r}
n <- estimateReconstructionParametersSPE(
    sostaSPE,
    marks = "cell_type",
    image_col = "image_name",
    mark_select = "A",
    plot_hist = TRUE
)
```
We will use the mean of the two estimated vectors as our parameters.

```{r}
(thres_spe <- mean(n$thres))
(bndw_spe <- mean(n$bndw))
```

We cam now use the function `reconstructShapeDensity` to segment the islet of one image. The result is a `r BiocStyle::CRANpkg('sf')` polygon [@pebesmaSimpleFeaturesStandardized2018]. 

```{r}
struct <- reconstructShapeDensityImage(
    sostaSPE,
    marks = "cell_type",
    image_col = "image_name",
    image_id = "image1",
    mark_select = "A",
    bndw = bndw_spe,
    dim = 500,
    thres = thres_spe
)
```

We can plot both the points and the estimated islets polygon.

```{r}
plotSpots(
    sostaSPE[, sostaSPE[["image_name"]] %in% c("image1")],
    annotate = "cell_type",
    sample_id = "sample_id",
    in_tissue = NULL,
    y_reverse = FALSE,
) +
    geom_sf(
        data = struct,
        fill = NA,
        color = "darkblue",
        inherit.aes = FALSE, # this is important
        linewidth = 0.75
    )
```

If no arguments are given, the function `reconstructShapeDensityImage` estimates the reconstruction parameters internally.

```{r}
struct2 <- reconstructShapeDensityImage(
    sostaSPE,
    marks = "cell_type",
    image_col = "image_name",
    image_id = "image1",
    mark_select = "A",
    dim = 500
)
```


```{r}
plotSpots(
    sostaSPE[, sostaSPE[["image_name"]] %in% c("image1")],
    annotate = "cell_type",
    sample_id = "sample_id",
    in_tissue = NULL,
    y_reverse = FALSE,
) +
    geom_sf(
        data = struct2,
        fill = NA,
        color = "darkblue",
        inherit.aes = FALSE, # this is important
        linewidth = 0.75
    )
```

## Reconstruction of structures for all images

The function `reconstructShapeDensitySPE` reconstructs the islets for all images in the `spe` object. We use the estimated parameters from above.

```{r, eval=TRUE}
all_structs <- reconstructShapeDensitySPE(
    sostaSPE,
    marks = "cell_type",
    image_col = "image_name",
    mark_select = "A",
    bndw = bndw_spe,
    thres = thres_spe,
    ncores = 1
)
```

# Instersection with cells

Here we will assign cells in the `SingleCellExperiment` object to the specific islet.

```{r}
assign <- assingCellsToStructures(sostaSPE, all_structs, image_col = "image_name", n_cores = 8)
colData(sostaSPE)$struct_assign <- assign
```


```{r}
plotSpots(
    sostaSPE,
    annotate = "struct_assign",
    in_tissue = NULL,
    y_reverse = FALSE
) + facet_wrap(~image_name)
```
# Border cells

```{r}
st_all_structs
```


# Distance to cell types

```{r}

all_structures








computeDistances <- function(spe, struct_column, cell_type_column) {
    # Extract structure assignments from column in SPE
    structs <- unique(spe[[struct_column]])
    # Remove NA values
    structs <- structs[!is.na(structs)]
    # Compute the proportion of each cell type within each structure
    res <- lapply(structs, function(sel) {
        sub_spe <- sostaSPE[, sostaSPE[["image_name"]] %in% sel]
        sub_struct <- all_structures[all_structures[["image_name"]] %in% sel,]
        dist <- st_distance(spatialCoords2SF(sub_spe), sub_struct)
        res <- apply(dist, 1, min)

    })
    # Combine into single df and name with structs id
    res_mat <- do.call(rbind, res) |> as.data.frame()
    rownames(res_mat) <- structs
    return(res_mat)
}

```

# Distance to cell types

```{r}

```

