---
title: "Reconstruction and quantification of anatomical structures"
author: 
  - name: "Samuel Gunz"
    affiliation:
      - &DMLS Department of Molecular Life Sciences, University of Zurich, Switzerland
      - &SIB SIB Swiss Institute of Bioinformatics, University of Zurich, Switzerland
    email: "samuel.gunz@uzh.ch"
  - name: Mark D. Robinson
    affiliation:
      - *DMLS
      - *SIB
package: "`r BiocStyle::Biocpkg('sosta')`"
output:
  BiocStyle::html_document
abstract: >
  Reconstruction and analysis of anatimical structures using the `sosta` package
vignette: >
  %\VignetteIndexEntry{ Reconstruction and analysis of anatomical structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: sosta.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Installation

`r BiocStyle::Biocpkg('sosta')` can be loaded from Bioconductor and installed as follows

```{r installation, include = TRUE, eval=FALSE}
if (!requireNamespace("BiocManager")) {
    install.packages("BiocManager")
}
BiocManager::install("sosta")
```

# Setup

```{r setup, message=FALSE}
library("sosta")
library("dplyr")
library("tidyr")
library("ggplot2")
library("ggspavis")
library("sf")
library("SpatialExperiment")

theme_set(theme_bw())
```

# Introduction

As an example, we will load an simulated dataset with three different images and three cell types A, B and C.

```{r loading, echo=FALSE, message=FALSE}
# load the data
data("sostaSPE")
``` 

```{r}
plotSpots(
    sostaSPE,
    annotate = "cellType",
    sample_id = "sample_id",
    in_tissue = NULL,
    y_reverse = FALSE
) + facet_wrap(~imageName)
```

The goal is to reconstruct / segment the structures given by cell type A.

# Structure reconstruction

## Reconstruction of structures for one image

In this example we will reconstruct the islets based on the point pattern density of the cells of type A. We will start with estimating the parameters that we use for reconstruction afterwards. For one image this can be illustrated as follows.

```{r}
shapeIntensityImage(
    sostaSPE,
    marks = "cellType",
    imageCol = "imageName",
    imageId = "image1",
    markSelect = "A"
)
```

We see the density (pixel) image on the left and a histogram of the intensity values on the right. The estimated threshold is roughly the mean between the two modes of the (truncated) pixel intensity distribution.

This was done for one image. The function `estimateReconstructionParametersSPE` returns two plots with the estimated parameters for each image. The parameter `bndw` is the bandwidth parameter that is used for estimating the intensity profile of the point pattern. The parameter `thres` is the estimated parameter for the density threshold for reconstruction.

```{r}
n <- estimateReconstructionParametersSPE(
    sostaSPE,
    marks = "cellType",
    imageCol = "imageName",
    markSelect = "A",
    plotHist = TRUE
)
```
We will use the mean of the two estimated vectors as our parameters.

```{r}
(thresSPE <- mean(n$thres))
(bndwSPE <- mean(n$bndw))
```

We cam now use the function `reconstructShapeDensity` to segment the islet of one image. The result is a `r BiocStyle::CRANpkg('sf')` polygon [@pebesmaSimpleFeaturesStandardized2018]. 

```{r}
struct <- reconstructShapeDensityImage(
    sostaSPE,
    marks = "cellType",
    imageCol = "imageName",
    imageId = "image1",
    markSelect = "A",
    bndw = bndwSPE,
    dim = 500,
    thres = thresSPE
)
```

We can plot both the points and the segmented polygon.

```{r}
plotSpots(
    sostaSPE[, sostaSPE[["imageName"]] %in% c("image1")],
    annotate = "cellType",
    sample_id = "sample_id",
    in_tissue = NULL,
    y_reverse = FALSE,
) +
    geom_sf(
        data = struct,
        fill = NA,
        color = "darkblue",
        inherit.aes = FALSE, # this is important
        linewidth = 0.75
    )
```

If no arguments are given, the function `reconstructShapeDensityImage` estimates the reconstruction parameters internally.

```{r}
struct2 <- reconstructShapeDensityImage(
    sostaSPE,
    marks = "cellType",
    imageCol = "imageName",
    imageId = "image1",
    markSelect = "A",
    dim = 500
)
```


```{r}
plotSpots(
    sostaSPE[, sostaSPE[["imageName"]] %in% c("image1")],
    annotate = "cellType",
    sample_id = "sample_id",
    in_tissue = NULL,
    y_reverse = FALSE,
) +
    geom_sf(
        data = struct2,
        fill = NA,
        color = "darkblue",
        inherit.aes = FALSE, # this is important
        linewidth = 0.75
    )
```

## Reconstruction of structures for all images

The function `reconstructShapeDensitySPE` reconstructs the islets for all images in the `spe` object. We use the estimated parameters from above.

```{r, eval=TRUE}
allStructs <- reconstructShapeDensitySPE(
    sostaSPE,
    marks = "cellType",
    imageCol = "imageName",
    markSelect = "A",
    bndw = bndwSPE,
    thres = thresSPE,
    ncores = 1
)
```

## Instersection with cells

Here we will assign cells in the `SingleCellExperiment` object to the specific structure. 

```{r}
assign <- assingCellsToStructures(sostaSPE, allStructs, imageCol = "imageName", nCores = 1)
colData(sostaSPE)$structAssign <- assign
```


```{r}
plotSpots(
    sostaSPE,
    annotate = "structAssign",
    in_tissue = NULL,
    y_reverse = FALSE
) + facet_wrap(~imageName)
```

# Cell level metrics

## Distance to structure border

Using the function `minBoundaryDistances` we can compute the distance to the border structure. Negative values indicate that the points lie inside the structure.

```{r}
colData(sostaSPE)$minDist <- minBoundaryDistances(
    sostaSPE, "imageName",
    "structAssign", allStructs
)
plotSpots(
    sostaSPE,
    annotate = "minDist",
    in_tissue = NULL,
    y_reverse = FALSE
) +
    scale_colour_gradient2() +
    geom_sf(
        data = allStructs,
        fill = NA,
        inherit.aes = FALSE
    ) +
    facet_wrap(~imageName)
```

This info can be used to define border cells by thresholding to a range of positive and negative values.

```{r}
idx <- colData(sostaSPE)$minDist > -3 & colData(sostaSPE)$minDist < 3

colData(sostaSPE)$boder <- FALSE
colData(sostaSPE[, idx])$boder <- TRUE

plotSpots(
    sostaSPE,
    annotate = "boder",
    in_tissue = NULL,
    y_reverse = FALSE
) + geom_sf(
    data = allStructs,
    fill = NA,
    inherit.aes = FALSE
) +
    facet_wrap(~imageName)
```

Alternatively borders can be defined using `st_difference` and `st_buffer` on the structure object. In this case we will have an `sf` polygon that correspond to the border region.

```{r}
borders <- lapply(
    st_geometry(allStructs),
    \(x) st_difference(st_buffer(x, 3), st_buffer(x, -3))
) |>
    # both needed for proper conversion
    st_as_sfc() |> st_as_sf()

borders$imageName <- allStructs$imageName
borders$borderID <- paste0("border_", allStructs$structID)

colData(sostaSPE)$borderSf <- assingCellsToStructures(sostaSPE,
    borders,
    imageCol = "imageName",
    uniqueId = "borderID",
    nCores = 1
)
plotSpots(
    sostaSPE,
    annotate = "borderSf",
    in_tissue = NULL,
    y_reverse = FALSE
) + geom_sf(
    data = borders,
    fill = NA,
    inherit.aes = FALSE
) +
    facet_wrap(~imageName)
```

# Structure level metrics

## Proportion of cell types within structures

Using the function `cellTypeProportions` we can estimate the proportion of cell types witin each individual structure.

```{r}
cellTypeProportions(sostaSPE, "structAssign", "cellType")
```

## Shape Metrics


The function `totalShapeMetrics` calculates a set of metrics related to the shape of the structures

```{r}
shapeMetrics <- totalShapeMetrics(allStructs)
shapeMetrics

cbind(allStructs, t(shapeMetrics)) |>
    ggplot() +
    geom_sf(aes(fill = Area)) +
    facet_wrap(~imageName)
```

# Session Info

```{r sessionInfo}
sessionInfo()
```

# References
